HW3 Biost 578, Caitlin McHugh
========================================================


I initially install the required Bioconductor packages.
```{r,eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(c("GEOmetadb", "GEOquery"))
```

After installation, I load the required libraries.
```{r}
library(GEOmetadb); library(GEOquery)
library(Biobase)
```

I query the database to find the samples from the paper we desire: Henn, A. D. et al. High-resolution temporal response patterns to influenza vaccine reveal a distinct human plasma cell gene signature. Scientific Reports 3, 2327 (2013). I use the GEO accession number to access the exact samples we want to get; it's GSE29617. We download the raw data for this study.
```{r query-GEO, cache = TRUE}
# Download the mapping information and processed data
# This returns a list of eSets
GSE_set <- getGEO("GSE29617")[[1]]
dim(GSE_set) # 54715 80
phenoData(GSE_set)
dim(fData(GSE_set)) # 54715 16
```

We want to parse out the pertinent information from the phenoData for this study. First, we parse out the sample id and timepoints and store them in the phenoData as `subject' and `time.'
```{r}
names(pData(GSE_set))
table(pData(GSE_set)$characteristics_ch1.1) # has subject ID
table(pData(GSE_set)$characteristics_ch1.2) # has time point info

pData(GSE_set)$subject <- gsub("subject id: ","",pData(GSE_set)$characteristics_ch1.1)
pData(GSE_set)$time <- gsub("time point: ","",pData(GSE_set)$characteristics_ch1.2)

names(pData(GSE_set)) # good
```

To run voom/limma, we first create an eSet we can use, including the phenoData from our GSE dataset.
```{r}
library(limma)
# Note that I add one to the count
new_set <- ExpressionSet(assayData = exprs(GSE_set)+1)
pData(new_set) <- pData(GSE_set)
```

We now need to set-up our design matrix to estimate our weights, which includes variables for subject and the timepoints.
```{r}
design <- model.matrix(~subject+time, new_set)
new_set_voom <- voom(new_set,design = design)
```

Finally, we are able to fit the model and extract the baseline and day 3 comparison, as well as the comparison of baseline with day 7.
```{r}
lm <- lmFit(new_set_voom, design)
eb <- eBayes(lm)
# Look at the other time-points
t3 <- topTable(eb, coef = "timeD3", number = Inf)
t7 <- topTable(eb,coef="timeD7",number=Inf)

sum(t3$adj.P.Val<0.1) # 1392
sum(t7$adj.P.Val<0.1) # 869
sum(t3$adj.P.Val<0.1&t7$adj.P.Val<0.1) # 869
```

Now we need to make the heat map plot of these differentially expressed genes with an adjusted p-value of less than 0.1. We group the results by time point.
```{r}
library(pheatmap)
PValue <- matrix(c(t3$logFC[t3$adj.P.Val<0.1],t7$logFC[t3$adj.P.Val<0.1]),ncol=2)
rownames(PValue) <- rownames(t3)[t3$adj.P.Val<0.1]
colnames(PValue) <- c("timeD3","timeD7")
```

We are finally able to display the results in a heatmap, grouping by time point.
```{r fig.width=7, fig.height=6}
pheatmap(PValue,cluster_cols=FALSE,color=colorRampPalette(c("red","white","blue"))(50))
```

Let's first download and install the package that we need for the GSEA analysis.
```{r, eval=FALSE}
library(BiocInstaller)
biocLite("GSEABase")
```

We load the `GSEAbase` package for loading gene sets and convert the gene sets to gene indices.
```{r}
library(GSEABase)
c2_set <- getGmt("msigdb.v4.0.symbols.gmt.txt")
gene_ids <- geneIds(c2_set)
# Camera requires gene-indices
sets_indices <- symbols2indices(gene_ids, fData(GSE_set)[,"Gene Symbol"])
```

As with `limma`, we need to specify the contrast we wish to test at the set level.
```{r}
# Note that camera works on voom objects
res <- vector("list",length = 2)
ct <- 1
for(i in c(3,7))
  {
  contrast <- paste0("timeD",i)
  cont_matrix <- makeContrasts(contrast, levels=design)
  res[[ct]] <- camera(new_set_voom, sets_indices, design=design, contrast=cont_matrix, sort=FALSE)
  ct <- ct+1
}
```

We set up the results for plotting with the pheatmap function.
```{r}
library(pheatmap)
PValue <- sapply(res, function(x){ifelse(x$Direction=="Up", -10*log10(x$PValue), 10*log10(x$PValue))})
rownames(PValue) <- rownames(res[[1]])
PValue_max <- rowMax(abs(PValue))
PValue_small <- PValue[PValue_max>30, ]
anno <- data.frame(Time=paste0("timeD",c(3,7)))
rownames(anno) <- colnames(PValue_small)  <- paste0("timeD",c(3,7))
```

We are finally able to display the results in a heatmap, grouping by time point.
```{r fig.width=7, fig.height=6}
pheatmap(PValue_small, cluster_cols=FALSE,show_rownames=FALSE)
```